---
title: "Supplemental Information 2"
output: pdf_document
---

```{r setup, include=FALSE}
options(width = 300)
```

# Description

An R markdown of the phenotypic analysis and relevant information pertaining to interpretation of results. Analysis will be performed and results will be explained after relevant chunk information.  

```{r, echo=FALSE, results='hide'}
# Load necessary libraries
library(gaston)
library(tidyverse)
library(asreml)
library(ASMap)
library(qtl)
library(psych)
library(knitr)
library(ggplot2)
library(patchwork)

# Read in data
pheno <- read.csv("NC13955_Combined_Data_Final.csv")

# Read in potential selfs
selfs <- read.csv("list_of_potential_selfs.csv")

# Remove selfs
pheno <- pheno %>%
  filter(!GENOTYPE %in% selfs$GENOTYPE)

# Ensure the correct data types
pheno[, 1:7] <- lapply(pheno[, 1:7], as.factor)
pheno[, 8:ncol(pheno)] <- lapply(pheno[, 8:ncol(pheno)], as.numeric)

# Colnames
colnames(pheno)[8:ncol(pheno)] <- c("Heading Date (Days)",
                                    "Visual Rating (0-9)",
                                    "Fusarium Damaged Kernels (%)",
                                    "DON Content (PPM)", 
                                    "Plant Height (cm)")
```

# Single environment analysis

Here I am going through each environment and assessing the significance of the genotype effect using the following model 

$$
y_{ij}=\mu+G_i+r_j+\varepsilon_{ij}
$$

Where G is the fixed genotype effect, r is the random replication effect, and $\varepsilon$ is the residual error. Each loop will be denoted using "#" as the dilimeter and the environment assessed will be denoted by the assigned short hand (KIN = Kinston, NC; RAL = Raleigh, NC; WAR = Warsaw, VA; 19 = 2019; 20 = 2020). We will use a fully random model to estimate heritability via the following equation:

$$
h^2=\frac{\sigma^2_g}{\sigma^2_g+\sigma^2_\varepsilon}
$$

```{r}
# Set up lists for loop
env <- levels(pheno$ENV)
traits <- colnames(pheno)[8:12]

# Create dataframe for means output
BLUEs_loc <- data.frame(GENOTYPE = unique(pheno$GENOTYPE))
h2_envs <- c()

# Run models for each environment
for (i in env) {
  for (j in traits) {
    # Print message
    print_message <- paste("### Analyzing", j, "in", i, "###")
    print(paste(rep("#", nchar(print_message)), collapse = ""))
    print(print_message)
    print(paste(rep("#", nchar(print_message)), collapse = ""))
    
    # Read subset of phenotype data
    pheno_subset <- na.omit(pheno[pheno$ENV == i, c("GENOTYPE", "REP", j)])
    colnames(pheno_subset)[3] <- "Y"
    
    # Check if subset is empty
    if (nrow(na.omit(pheno_subset)) == 0) {
      # Throw message
      print(paste("###", j, "not taken in", i, "moving on..."))
      remove(pheno_subset)
      next
    }
    
    # If the trait is abnormally distributed
    if (j %in% c("Visual Rating (0-9)", "Fusarium Damaged Kernels (%)", "DON Content (PPM)")) {
      # Run model and pull estimates
      fit1 <-  suppressMessages(
        suppressWarnings(
          asreml(fixed = Y ~ GENOTYPE,
                 random = ~ REP,
                 residual = ~ units,
                 data = pheno_subset,
                 family = asr_poisson(),
                 maxit = 75,
                 trace = FALSE)
          )
        )
      
            # Run model and pull estimates
      fit2 <-  suppressMessages(
        suppressWarnings(
          asreml(fixed = Y ~ 1,
                 random = ~ GENOTYPE+REP,
                 residual = ~ units,
                 data = pheno_subset,
                 family = asr_poisson(),
                 maxit = 75,
                 trace = FALSE)
          )
        )
      
      # Check if model converges and if it results in critical error, 
      # default to RCBD with normal distribution
      if (fit1$converge == FALSE | (i == "WAR20" & j == "Visual Rating (0-9)")) {
        # Fit model
        fit1 <- suppressMessages(
          suppressWarnings(
            asreml(fixed = Y ~ GENOTYPE,
                   random = ~ REP,
                   residual = ~ units,
                   data = pheno_subset,
                   maxit = 75, 
                   trace = FALSE)
          )
        )
        
        # Fit model
        fit2 <- suppressMessages(
          suppressWarnings(
            asreml(fixed = Y ~ 1,
                   random = ~ GENOTYPE+REP,
                   residual = ~ units,
                   data = pheno_subset,
                   maxit = 75, 
                   trace = FALSE)
          )
        )
        # Update model
        fit1 <- update.asreml(fit1)
        fit2 <- update.asreml(fit2)
      }
      
      # Print message
      print_message <- paste("### Summary of fixed effect model for", j, "in", i, "###")
      print(paste(rep("#", nchar(print_message)), collapse = ""))
      print(print_message)
      print(paste(rep("#", nchar(print_message)), collapse = ""))
      
      # Get quick summary (comment out for markdown) 
      print(summary(fit1))
      
      # Pull BLUEs
      pred <- predict(fit1, classify = "GENOTYPE", maxit = 75)$pvals[, c("GENOTYPE", "predicted.value")]
      
      # Rename
      colnames(pred)[2] <- paste(j, "_", i, sep = "")
      
      # Bind
      BLUEs_loc <- left_join(BLUEs_loc, pred, by = "GENOTYPE")
      
      # Print message
      print_message <- paste("### Summary of random effect model for", j, "in", i, "###")
      print(paste(rep("#", nchar(print_message)), collapse = ""))
      print(print_message)
      print(paste(rep("#", nchar(print_message)), collapse = ""))
      
      # Print summary
      print(summary(fit2))
      
      # Estimate heritability
      h2 <- vpredict(fit2, h2~V2/(V2+V3))
      
      # Print message
      print_message <- paste("### Narrow-sense heritability of", j, "in", i, "###")
      print(paste(rep("#", nchar(print_message)), collapse = ""))
      print(print_message)
      print(paste(rep("#", nchar(print_message)), collapse = "")) 
      
      # Table
      print(knitr::kable(h2, digits = 2))
      
      # Add to h2
      h2_envs <- rbind(h2_envs,
                       data.frame(Environment = i,
                                  Trait = j,
                                  h2))
      
    } else {
      # Run model and pull estimates 
      fit1 <- suppressMessages(
        suppressWarnings(
          asreml(fixed = Y ~ GENOTYPE,
                 random = ~ REP,
                 residual = ~ units,
                 data = pheno_subset,
                 trace = FALSE)          
        )
      ) 
      
      # Run model and pull estimates 
      fit2 <- suppressMessages(
        suppressWarnings(
          asreml(fixed = Y ~ 1,
                 random = ~ GENOTYPE+REP,
                 residual = ~ units,
                 data = pheno_subset,
                 trace = FALSE)          
        )
      ) 
      
      # Print message
      print_message <- paste("### Summary of fixed effect model for", j, "in", i, "###")
      print(paste(rep("#", nchar(print_message)), collapse = ""))
      print(print_message)
      print(paste(rep("#", nchar(print_message)), collapse = ""))
      
      # Get quick summary (comment out for markdown) 
      print(summary(fit1))

      # Get BLUEs
      pred <- predict(fit1, classify = "GENOTYPE")$pvals[, c("GENOTYPE", "predicted.value")]
      
      # Rename
      colnames(pred)[2] <- paste(j, "_", i, sep = "")
      
      # Bind
      BLUEs_loc <- left_join(BLUEs_loc, pred, by = "GENOTYPE")
      
      # Print message
      print_message <- paste("### Summary of random effect model for", j, "in", i, "###")
      print(paste(rep("#", nchar(print_message)), collapse = ""))
      print(print_message)
      print(paste(rep("#", nchar(print_message)), collapse = ""))
      
      # Print summary
      print(summary(fit2))
      
      # Estimate heritability
      h2 <- vpredict(fit2, h2~V2/(V2+V3))
      
      # Print message
      print_message <- paste("### Narrow-sense heritability of", j, "in", i, "###")
      print(paste(rep("#", nchar(print_message)), collapse = ""))
      print(print_message)
      print(paste(rep("#", nchar(print_message)), collapse = "")) 
      
      # Table
      print(knitr::kable(h2, digits = 2))
      
      # Add to h2
      h2_envs <- rbind(h2_envs,
                       data.frame(Environment = i,
                                  Trait = j,
                                  h2))      
      
    }
    # Remove temporary objects
    remove(pheno_subset, fit1, fit2, pred, h2)
  }
}
```

## Visualization

Here I visualize the results of the above loops. Output from these loops are pairs plots which show scatter plots fit with a linear regression in the lower triangle. The diagonal of these plots shows the distribution of each trait which is listed as plain text above the distribution with the corresponding name of the environment (KIN = Kinston, NC; RAL = Raleigh, NC; WAR = Warsaw, VA; 19 = 2019; 20 = 2020).

```{r, fig.width=10, fig.height=10}
# For i in environments and traits
for(i in traits) {
  # Parse blues and h2
  i <- gsub("\\([^\\)]*\\)", "", i)
  i <- trimws(i)
  temp1 <- BLUEs_loc[,grep(i,colnames(BLUEs_loc))]
  temp2 <- h2_envs[grep(i, h2_envs$Trait),]
  colnames(temp1) <- gsub("_", " ",colnames(temp1))
  
  pairs.panels(temp1,
               hist.col = "gray",
               lm = TRUE,
               stars = TRUE,
               digits = 2,
               density = FALSE,
               ellipses = FALSE,
               main = paste("Visualization of", i, "correlations across environments"))
}
```

# Assess Genotype-by-Environment Analysis

Here I run a model for each trait:

$$
y_{ijk}=\mu+G_i+E_j+GE_{ij}+er_{jk}+\varepsilon_{ijk}
$$

Where G is the fixed genotype effect, E is the fixed environment effect, GE is the fixed genotype by environment interaction, er is the random environment by replication interaction, and $\varepsilon$ is the residual error. All random effects are individually and identically distributed. 

```{r}
# Define traits
traits <- colnames(pheno)[8:12]

# Create dataframe for predictions
BLUEs_mem <- data.frame(GENOTYPE = unique(pheno$GENOTYPE))

# Run MEMLM loop for each trait
for (i in traits) {
  
  # Print message for analyzing each trait
  print_message <- paste("### Analyzing", i, "###")
  print(paste(rep("#", nchar(print_message)), collapse = ""))
  print(print_message)
  print(paste(rep("#", nchar(print_message)), collapse = ""))
  
  # Subset data
  pheno_subset <- na.omit(pheno[, c("GENOTYPE", "YEAR", "ENV", "REP", i)])
  
  # Redefine data types
  pheno_subset[, 1:4] <- lapply(pheno_subset[, 1:4], as.factor)
  pheno_subset[, 5] <- as.numeric(pheno_subset[, 5])
  colnames(pheno_subset)[5] <- "Y"
  
  # Run mixed effects model for traits
  if (i %in% c("Visual Rating (0-9)", "Fusarium Damaged Kernels (%)", "DON Content (PPM)")) {
    # Suppress output from the first model with Poisson distribution
    suppressWarnings(suppressMessages({
      capture.output({
        fit <- asreml(fixed = Y ~ GENOTYPE,
                      random = ~ ENV + GENOTYPE:ENV + REP:ENV,
                      residual = ~ units,
                      data = pheno_subset,
                      family = asr_poisson(),
                      maxit = 75,
                      trace = FALSE)
      }, file = NULL)
    }))
    
    # Suppress output from the second model with Poisson distribution
    suppressWarnings(suppressMessages({
      capture.output({
        fit2 <- asreml(fixed = Y ~ GENOTYPE + ENV + GENOTYPE:ENV,
                       random = ~ REP:ENV,
                       residual = ~ units,
                       data = pheno_subset,
                       family = asr_poisson(),
                       maxit = 75,
                       trace = FALSE)
      }, file = NULL)
    }))
    
    # If model fails to converge, fit normal distribution model and suppress output
    if(fit$converge==FALSE | fit2$converge==FALSE){
      suppressWarnings(suppressMessages({
        capture.output({
          fit <- asreml(fixed = Y ~ GENOTYPE,
                        random = ~ ENV + GENOTYPE:ENV + REP:ENV,
                        residual = ~ units,
                        data = pheno_subset,
                        maxit = 75,
                        trace = FALSE)
        }, file = NULL)
      }))
      
      suppressWarnings(suppressMessages({
        capture.output({
          fit2 <- asreml(fixed = Y ~ GENOTYPE + ENV + GENOTYPE:ENV ,
                         random = ~ REP:ENV,
                         residual = ~ units,
                         data = pheno_subset,
                         maxit = 75,
                         trace = FALSE)
        }, file = NULL)
      }))
    }
  } else {
    # Suppress output from the model with normal distribution
    suppressWarnings(suppressMessages({
      capture.output({
        fit <- asreml(fixed = Y ~ GENOTYPE,
                      random = ~ ENV + GENOTYPE:ENV + REP:ENV,
                      residual = ~ units,
                      data = pheno_subset,
                      maxit = 75,
                      trace = FALSE)
      }, file = NULL)
    }))
    
    # Suppress output from the second model with normal distribution
    suppressWarnings(suppressMessages({
      capture.output({
        fit2 <- asreml(fixed = Y ~ GENOTYPE + ENV + GENOTYPE:ENV ,
                       random = ~ REP:ENV,
                       residual = ~ units,
                       data = pheno_subset,
                       maxit = 75,
                       trace = FALSE)
      }, file = NULL)
    }))
  }
  
  # Print summary of Wald test for fit2
  temp1 <- capture.output({
    wald_output <- as.data.frame(wald(fit2))
  }, file = NULL)
  
  print(wald_output)
  
  # Get predictions for GENOTYPE
  pred <- predict(fit, classify = "GENOTYPE")$pvals[, 1:2]
  colnames(pred)[2] <- paste(i, "ME", sep = "_") 
  
  # Left join predictions to the main dataframe
  BLUEs_mem <- left_join(BLUEs_mem, pred, by = "GENOTYPE")
  
  # Remove temporary objects
  rm(pheno_subset, pred)
}
```

# Heritability Variance Component Estimates

Here we will us a fully random version of the above model to assess the per-plot and entry-mean heritability of each trait across environments. Variance components will be shown for each loop. 

```{r}
# Make an empty data frame for the heritability estimates
h2s <- data.frame(Trait = character(), 
                  Type = character(), 
                  Estimation = numeric(), 
                  SE = numeric())

# Calculate the harmonic means for each trait
e_traits <- list()
r_traits <- list()

for (j in traits) {
  # Subset data and only take complete cases
  a <- pheno %>% 
    drop_na(all_of(j))
  
  # Get the number of observations per genotype
  obs_per_ind <- aggregate(a[, j] ~ GENOTYPE, 
                           data = a,
                           length)
  
  # Calculate the harmonic mean of replications
  r <- (1 / mean(1 / obs_per_ind$`a[, j]`))
  
  # Store harmonic mean in r_traits list
  r_traits[j] <- r
  
  # Create a new object called e
  e <- c()
  
  # For each genotype in the phenotypic dataset
  for (i in unique(pheno$GENOTYPE)) {
    # Get all observations for each genotype
    a <- na.omit(pheno[pheno$GENOTYPE == i, c("ENV", j)])
    
    # Get the number of environments the genotype was observed in
    a <- length(unique(a$ENV))
    
    # Calculate harmonic mean for environments
    a <- ifelse(a == 0, 0, 1 / a)
    
    # Add this value to e
    e <- rbind(e, a)
  }
  
  # Store harmonic mean of environments in e_traits list
  e_traits[j] <- (1 / mean(e)) 
  
  # Clean up temporary variables
  remove(a, obs_per_ind, r, e)
}

# Loop through each trait for heritability estimation
for (i in traits) {
  # Print message for analyzing each trait
  print_message <- paste("### Analyzing", i, "###")
  print(paste(rep("#", nchar(print_message)), collapse = ""))
  print(print_message)
  print(paste(rep("#", nchar(print_message)), collapse = ""))
  
  # Determine number of locations
  n_loc <- pheno %>% 
    select(ENV, YEAR, LOC, all_of(i)) %>% 
    drop_na() %>% 
    distinct(ENV)
  n_loc <- as.numeric(nrow(n_loc))
  
  # Determine the number of replications
  n_rep <- n_loc * 2
  
  # Determine the number of years
  n_year <- pheno %>% 
    select(ENV, YEAR, LOC, all_of(i)) %>% 
    drop_na() %>% 
    distinct(YEAR)
  n_year <- as.numeric(count(n_year))
  
  # Get the harmonic mean of environments
  e <- as.numeric(e_traits[i])
  
  # Get the harmonic mean of replications
  r <- as.numeric(r_traits[i])
  
  # If the trait has been observed over many years
  if (n_year >= 2) {
    # If the trait is a disease reaction trait
    if (i %in% c("Visual Rating (0-9)", "Fusarium Damaged Kernels (%)", "DON Content (PPM)")) {
      # Run model for Poisson distributed traits
      mlm <- asreml(fixed = pheno[, i] ~ 1,
                    random = ~ GENOTYPE +
                      ENV +
                      GENOTYPE:ENV +
                      REP:ENV,
                    residual = ~ idv(units),
                    data = pheno,
                    maxit = 75,
                    family = asr_poisson(),
                    trace = FALSE)
    } else {
      # Run model for normally distributed traits
      mlm <- asreml(fixed = pheno[, i] ~ 1,
                    random = ~ (GENOTYPE) +
                      ENV +
                      GENOTYPE:ENV+
                      REP:ENV,
                    residual = ~ idv(units),
                    data = pheno,
                    maxit = 75,
                    trace = FALSE)
    }
  # If the trait has been only observed in one year
  } else {
    # Run model for single year data
    mlm <- asreml(fixed = pheno[, i] ~ 1,
                  random = ~ (GENOTYPE) +
                    ENV +
                    GENOTYPE:ENV+
                    REP:ENV,
                  residual = ~ idv(units),
                  data = pheno,
                  maxit = 75,
                  trace = FALSE)
  }
  
  # Print summary of variance components
  print(summary(mlm)$varcomp)
  
  # Predict heritability estimates
  pph2 <- vpredict(mlm, 
                   h2 ~ V2 / (V2 + V3 + V4))
  
  emh2 <- vpredict(mlm,
                   h2 ~ V2 / (V2 + (V3 / e) + (V4 / (e * r))))
  
  # Bind predictions to h2s data frame
  a <- rbind(pph2, emh2)
  a <- cbind(data.frame(Trait = c(i, i)), data.frame(Type = c("Per-Plot", "Entry-Mean")), a)
  h2s <- rbind(h2s, a)
  
  # Clean up temporary variables
  remove(mlm, pph2, emh2, n_loc, n_rep, n_year, e, r, a)
}
```
